\vspace{-1.75cm}PhD Proposal

### 2022-05-15 abstract/abstract.tex
continuo abstract

\documentclass[a4paper,11pt, oneside]{article}
\usepackage[left=15mm, right=15mm, top=15mm, bottom=20mm]{geometry}
\usepackage[utf8]{inputenc}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{graphics}
\usepackage{amsfonts}
\usepackage{listings}
\usepackage{amsmath}
\usepackage{amstext}
\usepackage{engrec}
\usepackage{rotating}
\usepackage{verbatim}
\usepackage[safe,extra]{tipa}
% \usepackage{showkeys}
\usepackage{multirow}
\usepackage{hyperref}
\usepackage{microtype}
\usepackage{fontspec}
\usepackage{enumerate}
\usepackage{physics}
\usepackage{braket}
\usepackage{mhchem}
\usepackage{marginnote}
\usepackage{pgfplots}
\usepackage{cancel}
\usepackage{polynom}
\usepackage{booktabs}
\usepackage{enumitem}
\usepackage{framed}
\usepackage{pdfpages}
\usepackage{pgfplots}
\usepackage{algorithm}
% \usepackage{algpseudocode}
\usepackage[cache=false]{minted}
\usepackage{mathtools}
\usepackage[noend]{algpseudocode}

\newcommand*{\bfrac}[2]{\genfrac{}{}{0pt}{}{#1}{#2}}

\usepackage{tikz}\usetikzlibrary{er}\tikzset{multi  attribute /.style={attribute
    ,double  distance =1.5pt}}\tikzset{derived  attribute /.style={attribute
    ,dashed}}\tikzset{total /.style={double  distance =1.5pt}}\tikzset{every
  entity /.style={draw=orange , fill=orange!20}}\tikzset{every  attribute
  /.style={draw=MediumPurple1, fill=MediumPurple1!20}}\tikzset{every
  relationship /.style={draw=Chartreuse2,
    fill=Chartreuse2!20}}\newcommand{\key}[1]{\underline{#1}}
\usetikzlibrary{arrows.meta}
\usetikzlibrary{decorations.markings}
\usetikzlibrary{arrows,shapes,backgrounds,petri}
\tikzset{
  place/.style={
    circle,
    thick,
    draw=black,
    minimum size=6mm,
  },
  transition/.style={
    rectangle,
    thick,
    fill=black,
    minimum width=8mm,
    inner ysep=2pt
  },
  transitionv/.style={
    rectangle,
    thick,
    fill=black,
    minimum height=8mm,
    inner xsep=2pt
  }
} 
\usetikzlibrary{automata,positioning,chains,fit,shapes}
\usetikzlibrary{circuits.logic.US}
\usetikzlibrary{positioning}
\usepackage{fancyhdr}
% \pagestyle{fancy}
% \fancyhead[LE,RO]{\slshape \rightmark}
% \fancyhead[LO,RE]{\slshape \leftmark}
\fancyfoot[C]{\thepage}
\usepackage[usenames,dvipsnames]{pstricks}
\usepackage{epsfig}
\usepackage{pst-grad} % For gradients
\usepackage{pst-plot} % For axes
\usepackage[space]{grffile} % For spaces in paths
\usepackage{etoolbox} % For spaces in paths
\makeatletter % For spaces in paths
\patchcmd\Gread@eps{\@inputcheck#1 }{\@inputcheck"#1"\relax}{}{}
\makeatother
\usepackage{lipsum}
\DeclareSymbolFont{symbolsC}{U}{txsyc}{m}{n}
\DeclareMathSymbol{\strictif}{\mathrel}{symbolsC}{74}
\title{\vspace{-1.75cm}PhD Proposal}
\author{Davide Cozzi, 829827,
  \href{mailto:d.cozzi@campus.unimib.it}{d.cozzi@campus.unimib.it}} 
\date{}
\makeatletter
\renewcommand{\paragraph}{%
  \@startsection{paragraph}{4}%
  {\z@}{1.25ex \@plus 1ex \@minus .2ex}{-1em}%
  {\normalfont\normalsize\bfseries}%
}
\makeatother
\usepackage{setspace}
\begin{document}
\maketitle
\definecolor{shadecolor}{gray}{0.80}
\setlist{leftmargin = 2cm}
\newtheorem{teorema}{Teorema}
\newtheorem{definizione}{Definizione}
\newtheorem{esempio}{Esempio}
\newtheorem{corollario}{Corollario}
\newtheorem{lemma}{Lemma}
\newtheorem{osservazione}{Osservazione}
\newtheorem{nota}{Nota}
\newtheorem{esercizio}{Esercizio}
\algdef{SE}[DOWHILE]{Do}{doWhile}{\algorithmicdo}[1]{\algorithmicwhile\ #1}

\renewcommand{\sectionmark}[1]{\markright{\thesection.\ #1}}
\newcommand{\floor}[1]{\lfloor #1 \rfloor}
\newcommand{\MYhref}[3][blue]{\href{#2}{\color{#1}{#3}}}%

@@

\textit{Computational pangenomics} is becoming increasingly essential in the

@@

\textbf{genome-wide association studies (\textit{GWAS})}. Unfortunately, in
order to complete these types of studies a large amount of data is needed, data
that must be indexed, queried and analyzed. Just
to give some examples the \textit{homo sapiens reference genome (GRCh38.p14)},
has a size of $\sim 3.1$\textit{gb} and contains $\sim 59,265$ genes. From
a biological point of view, it is also interesting to note that, as it has been 
pointed out by the study of \textit{1000 Genome Project}, there are over 88

@@

are \textbf{Single 
  Nucleotide Polymorphisms (\textit{SNPs})}, 3.6 million are \textbf{short
  insertions/deletions (\textit{indel})} and 60000 are \textbf{structural
  variants}, involving more than 50 nucleotides. Moreover it is necessary to
consider that the objective is the sequencing of at least 100 thousand samples,
in the next few years. It is therefore clear that all these data are now a limit
to the state of art algorithms and data structures.\\
The problem of \textit{pattern matching} is one of the most studied topics in
the field of algorithmics and bioinformatics. The interest in such

@@

  transform (\textit{BWT})}, thanks to the studies of Ferragina and Manzini, who
proposed its use together with indexing via \textit{FM-index}. The use of such
algorithms, such as \textit{BTW}, is essential in alignment algorithms such as

@@

order to compute the alignment, increasing the efficiency of the algorithm.\\  

@@

in the field of bioinformatics. Until a few years ago the research was focused
on the study of a \textbf{linear sequence of a genome} while now the 
researchers are beginning to deepen the topic of the \textbf{pangenome}, which
term was 
introduced by Tettelin in 2005. briefly, the \textit{pangenome} is a compact
rappresentation of multiple genomes, encoding the variations in a multitude of

@@

\cite{pangenome}. \\

@@

  sequencing}, which had led both 
to reduce the costs of single sequencing and to produce sequences of ever higher
quality in less and less time, the researchers were able to theorize the
\textbf{pangenome graph}. Furthermore, the new amount of sequences has led to
new algorithms regarding the problem of \textit{pattern matching}. In 2021,
Rossi et al. proposed \textit{MONI} as a data structure to handle a 
\textbf{run-length encoded version of BWT (\textit{RLBWT})}, with the ultimate
intention of indexing and using multiple genomes as a reference
\cite{moni}. Together with this data structure, the authors proposed the concept
of \textbf{matching statistics (\textit{MS})} in order to efficiently compute
the matches between a pattern and a text. A recent improvement has been made
through the implementation of \textit{PHONI} \cite{phoni}, where the
\textbf{longest-common-extension (\textit{LCE}) queries} in order to further
optimize the pattern search.\\ 
Against this background, various algorithms and data structures have
been implemented in order to study \textit{haplotypes} and
\textit{genotypes}, such as the \textbf{genotyping variants
  problem}. Briefly, we could define \textit{haplotypes} as a combination 
of allelic variants, each one inherited from a parent. Instead, the
\textit{genotype} is the complete set of genes contained in the DNA. So, from
2005, publication of the \textbf{GWAS} has begun. The goal of this type of
studies is to screen the  

@@

example in order to study outocomes of diseases.
In this particular historical period, it is impossible not to mention
viruses. In fact, by nature, viruses replicate a lot but
often in a not perfect way, during infections. Thus produces many inexact
clones, referred as 
\textbf{viral haplotypes}, which, taken together, form the \textbf{viral
  pangenome}. The identification of all these haplotypes is crucial both to the
study of the spread of viruses and to the production of efficient
drugs, in a context of high pharmacological resistance. \\

@@

  (\textit{PBWT})}, proposed by Durbin in 2014 \cite{pbwt_durbin}. Using this
particular data structure (which will be described below), it is possible to
study efficiently a collection of haplotypes but only in the bi-allelic
case. Furthermore variants of the \textit{PBWT} have been studied for
handling the multiallelic case. The use of the \textit{PBWT} is found in many
software for the study of haplotypes and in various genotype imputation methods,
that are studies that infers unobserved genotypes in a sample of individuals
\cite{impute5}.\\
During the development of my master thesis, I worked to create a run length
encoded variant of the \textit{PBWT}, the \textbf{RLPBWT}, using and adapting
the various theories developed for the \textit{RLBWT}, in collaboration with the
authors of \textit{MONI} and \textit{PHONI}. In this context, my PhD
is going to be focused on the development of new algorithms in
various topics related to open problems in the study of the
\textit{haplotyping/genotyping}, of the \textit{genome 
  variants} and of the imputation issues related. My intention is also to deepen
experimental themes  
of \textit{pattern matching} and of the \textit{pangenome graph}, in detail the
new developments on \textit{BWT} and indexing structures, as well as
\textit{succinct data structures}, with particular attention to the use of
\textit{bitvectors}.

@@

Now I present a brief overview of the main algorithms, data structures,
methods etc $\ldots$ that will be the core of my studies during PhD. 

@@

\paragraph*{Bitvectors}
\textbf{Bitvectors} are one of the most important data structure when
mentioning \textit{succinct data structures}. 

@@

integer $i$, such that $0<i\leq rank_B(n)$, where $n=|B|$, we can define

@@

\textit{constant time}, with the additional cost of $\mathcal{O}(n)$ bits in
memory. In more practical terms, there are several implementations of the 
same within \textbf{SDSL (\textit{Succint Data Structures Library})}, one of the
most important C++ library used in bioinformatics. As 
the implementation changes (for example \textit{plain bitvector, interleaved
  bitvector, sparse bivector} etc$\ldots$) the computational time of the two
operations varies (usually only one of the two is in constant time) as well as
the amount of additional bits needed. An example of the use of bitvectors is

@@

along the runs themselves. 
\paragraph*{RLBWT}

@@

\textit{FM-index}.\\
Speaking of \textit{pangenome}, linear indexing via FM-index is no longer the

@@

the \textbf{Run-Length Burrows–Wheeler Transform (\textit{RLBWT})}. Given a text

@@

\textbf{thresholds} (minimum \textit{LCP} value between two consecutive runs of
the same character) in \textit{MONI}, and the use of \textbf{longest common
  extension (\textit{LCE}) query} (to compute the right equal common extension
between two position in the text) in \textit{PHONI}. Both solutions use
\textbf{straight-line programs (\textit{SLP})} briefly a
\textit{grammar-compression} algorithm based on a \textit{context-free grammar},
for \textit{random access} in \textit{MONI} and for \textit{Longest Common
  extensions (LCE) queries} in \textit{PHONI}. 

@@

extended either right or left. Furthermore, using a particular

@@

in $SA$, the previous and next position in the suffix array. \\

@@

Thanks to these and other methods, it was possible to perform pattern matching

@@

a pangenomic context. In fact, most studies in the field of bioinformatics start
with the resolution of pattern matching problems and, as a direct consequence,
of alignment problems. 
\paragraph*{PBWT}

@@

iff $h$ is the smallest column index such that $y_i^k[h,k)=y_{i-1}^k[h,k)$. \\

@@

research based on it, both in terms of variant design, such as the already
mentioned multiallelic version or the \textbf{dynamic PBWT (\textit{d-PBWT}}),
and in terms of its use to study haplotype panels.\\

@@

signatures of natural selection.\\
Another interesting paper is the one by Wlliams and Mumey, published in 2020,

@@

addition of \textit{missing data}, handled with the help of wildcards, using the
\textit{PBWT}. \\ 

@@

software is to make genotype imputation in order to predict unobserved genotypes
from a panel with milions of haplotypes. Due to the size of the panel, the use
of \textit{PBWT} has been proved to be inevitable, further demonstrating the
importance of this data structure.\\
Durbin himself, the author of the \textit{PBWT}, with Shchur et al. in 2019,
proposed a use of his data structure in \textit{GWAS} context. In fact, they
studied the associations between genetic variants in order to identify signals
of natural selection and to build the so-called \textbf{ancestral recombination
  graph}, that contains complete information about history of samples.\\
For the sake of completeness, an example where \textit{PBWT} is not used is
\textbf{Ranbow}, proposed by Moeinzadeh et al. in 2020. The 
aim of this project was the haplotype reconstruction of polyploid genome from
short read sequencing data, studying also the multi-allelic case.

@@

Therefore, for my master's thesis, I had to rethink the concept of
\textit{Matching Statistics} for \textit{PBWT} (tracking  a \textit{row} of the
panel instead of the \textit{pos}), how to compute the \textit{SLP}
for the panel, how to use \textit{thresholds}/\textit{LCE queries} and how to
get the same behaviour of the $\varphi$ function, in order to obtain the
\textbf{RLPBWT}, combining the ideas related to the \textit{RLBWT} with those of
the \textit{PBWT}, eventually, as for example for the $\varphi$ function,
developing a simple new data structure.

@@

context. Summarizing, the most important research goals of my PhD are:

@@

  current use of bitvectors to handle efficiently the \textit{LF-mapping},
  with more than two alleles, too. 

@@

  Handling missing data is known to be \textit{hard} so, I should
  probably deal with parametric or approximate algorithms, based on
  researches already developed in \textit{BIAS}.
  \item \textbf{Genotype imputation \& other goals}. As introduced, I have
  developed a wide-range interest in the application of computer science to
  biology. So, I do 
  not intend to forget the actual use of the data structures that I will study
  and develop during my PhD. The management of missing and
  multiallelic data can further improve the state of the art of \textit{GWAS}
  and of genotype imputation, allowing even more precise inference of unobserved
  genotypes. All this must be read in the perspective of 
  a continuous rise of available data, not only regarding the human
  sequencing. \\
  I would also  point out my interest in deepening the issues related to
  pattern matching, indexing structures and pangenome graph. During
  my PhD, it’s my intention to focus on all the BIAS laboratory's topics of
  interest, from the point of view of bioinformatics and experimental 
  algorithmics. 

@@

From a more technical point of view I mainly focused on the use of the
\textbf{C++ programming language}, during both my bachelor and master
thesis projects. I managed to do this because of the availability of efficient
libraries, as the already 
cited \textit{SDSL}, that has become a standard in bioinformatics. However, in

@@

\textit{computational systems biology}. \\  

@@

my master thesis project, including Christina Boucher (University of Florida)
and Travis Gagie (Dalhousie University), in order to improve the quality of my
PhD program.
\vspace{-4mm}
\bibliographystyle{unsrt}
\bibliography{abstract}

>>>>

\documentclass[a4paper,11pt, oneside]{article}
\usepackage[left=15mm, right=15mm, top=15mm, bottom=15mm]{geometry}
\usepackage[utf8]{inputenc}
\usepackage{hyperref}
\usepackage{framed}
\usepackage{xcolor}
\usepackage{enumitem}
\usepackage[tickmarkheight=0.7ex]{todonotes}
\newcommand{\gdv}[1]{\todo{\textbf{GDV} #1}}
\newcommand{\pb}[1]{\todo{\textbf{PB} #1}}
\newcommand{\dc}[1]{\todo{\textbf{DC} #1}}
\usepackage[style=numeric, sorting=none, doi=false, isbn=false, url=false,
eprint=false,maxcitenames=2, mincitenames=1, maxbibnames=2, minbibnames=2,
giveninits=true]{biblatex} 
\addbibresource{abstract.bib}

\usepackage{fancyhdr}
% \pagestyle{fancy}
% \fancyhead[LE,RO]{\slshape \rightmark}
% \fancyhead[LO,RE]{\slshape \leftmark}
\fancyfoot[C]{\thepage}

\title{\vspace{-1.75cm}PhD Proposal}
\author{Davide Cozzi, 829827,
  \href{mailto:d.cozzi@campus.unimib.it}{d.cozzi@campus.unimib.it}} 
\date{}
\makeatletter
\renewcommand{\paragraph}{%
  \@startsection{paragraph}{4}%
  {\z@}{0.75ex \@plus 1ex \@minus .2ex}{-1em}%
  {\normalfont\normalsize\bfseries}%
}
\makeatother

\begin{document}
\maketitle
\setlist{leftmargin = 2cm}

@@

\textit{Computational pangenomics} is a new emerging research field in
computational biology. This concept is based on the idea we need to move
from the traditional view of a reference genome as a \textit{linear sequence} to
the one where we consider the \textit{genetic variations} or \textit{variants},
in a large collection of sequences of species. This new reference is called a
\emph{pangenome}.  
Pangenomics is becoming increasingly essential in the

@@

\textbf{genome-wide association studies (\textit{GWAS})} since GWAS studies need
to analyze a large collection of individual genomes.  Unfortunately, in 
order to complete these types of studies on a large amount of data, such data
must be indexed, queried and analyzed. Just to give some examples the
\textit{homo sapiens reference genome (GRCh38.p14)}, 
has a size of $\sim 3.1$\textit{gb} and contains $\sim 59,265$ genes, as
reported by NCBI. Biological studies done with the \textit{1000 Genome Project}
have pointed out that there are over 88 

@@

are \textbf{Single Nucleotide Polymorphisms (\textit{SNPs})}, 3.6 million are
\textbf{short insertions/deletions (\textit{indel})} and 60000 are
\textbf{structural variants}, involving more than 50 nucleotides. Moreover it is
necessary to consider that the objective is the sequencing of at least 100
thousand individuals, in the next few years. Therefore, it is clear that dealing
with  all these data is challenging for the current state of art algorithms and
data structures for sequence analysis. 
An example is represented by a classical research topic in sequence comparison:
\textit{pattern matching}, where the input is a long text (a
genome) and a short pattern (a read) and the output is the list of all
occurrences of the pattern as substrings of the text.  
The problem of \textit{pattern matching} is one of the most
fundamental topics in 
the field of algorithmics and bioinformatics. We can notice that such problem
has linear solution in time, if there is a proper indexing of the text. The
interest in such 

@@

  transform (\textit{BWT})} together with indexing via \textit{FM-index}. Thanks
to these algorithms and to the fact that it is retunerd an interval over the
\textit{BWT}, the problem of pattern matching, over an indexed text, has become 
linear on the length of pattern. The
use of such 
algorithms is essential in speeding alignment algorithms such as

@@

order to compute the alignment, increasing the efficiency of the algorithm. The
key point is the ability to compress large texts and to query the compressed
texts themselves.\\  

@@

in the field of bioinformatics. Now the 
researchers are deepening the topic of the \textbf{pangenome}, which
term was introduced
% by Tettelin
in 2005 for dealing with the comparison of genes in
bacterial species. Briefly, the \textit{pangenome} is a compact 
representation of multiple genomes, encoding the variations in a multitude of

@@

% \cite{pangenome}
\cite{pancon}. \\
The main focus of this proposal is improving the current state of the art
of pattern matching algorithms and data structures to manage biological
sequences, with a focus on spatial 
complexity. In fact, the design of data structures to handle large
collections of data, filling the gap between the genomic data production and the
current computational state of the art, will have an impact on the ability of
molecular biologists to analyze the current amount of data, represented by
pangenome graphs. 
%\pb{the main focus of this project proposal is...}
%\pb{IO menzionerei il tempo di calcolo del pattern matching del pattern ad un
%testo che è lineare nel pattern se vi è una adeguata rappresentazione o
%indicizzazione del testo} OK
% \pb{il tempo che si vuole avere per ricercare tutte le occorrenze del pattern
% con la BWT è lineare nel pattern anche se il numero di occorrenze potrebbe
% essere più alto perchè si restituisce un intervallo...in altre parole tieni
% presente che dovresti dare motivazioni all'uso dell'FM-index, tenere compresso
% il testo e nel contempo avere tutte le occorrenze come intervallo di
% BWT... dire qualcosa tipo che la dimensione dei dati richieste tenere il dato
% in forma compressa e quindi poter fare le query direttamente sul dato
% compresso...puoi anche dire che lo sviluppo e il disegno di strutture succinte
% per trattare grandi collezioni di dati e la loro implementazione e
% applicazione a sequenze reali (esempio aplotipi) è uno dei temi focus del tuo
% progetto...} CIRCA OK
% \pb{tenuto conto che il focus è filling  the current gap between the genomic
% data production and the ability of computer scintests to manage such data by
% developing new  data structure and algorithms that have to deal with these
% data. This improvement will have an impact on the ability of molecolar
% biologists to analyze the current amount of data represented by pangenome
% graphs as I will discuss in the specific goals of my project...} OK
\subsubsection*{Preliminaries}

@@

  sequencing}, which had led both to reduce the costs of single sequencing and
to produce sequences of ever higher quality in less and less time, the
researchers were able to theorize the \textbf{pangenome graph}. Furthermore, the
new amount of sequences has led to new algorithms for \textit{pattern matching}
for collection of sequences. In 2021, Rossi et al. proposed \textit{MONI} as a
data structure to handle a \textbf{run-length encoded version of BWT
  (\textit{RLBWT})}, with the ultimate intention of indexing and using multiple
genomes as a reference \cite{moni}. With this type of encoding, it is possibile
to further improve the compression of the text, having that consecutive
identical characters, the so-called \textit{runs}, are compactly stored, making
it possible to study pangenomic sequences. According to Rossi et al.,
\textit{BWA-MEM}, one of the most used read aligners, uses between 1.1 and 3.8
times more memory than \textit{MONI}. Together with this data structure, the
authors used the concept of \textbf{matching statistics (\textit{MS})} in order
to efficiently compute the matches between a pattern and a text. A recent
improvement has been made through the implementation of \textit{PHONI}
\cite{phoni}, where the \textbf{longest-common-extension (\textit{LCE})
  queries}, achieving the same results of \textit{MONI} with only a single sweep
over \textit{MS} array instead of two as in MONI (details in the next section).
On the other side, several computational problems arising from computational
biology has led to the development and implementation of algorithms and data
structures for analyzing specific sequence data such as \textit{haplotypes} and
\textit{genotypes} in the \textbf{genotyping variants problem}. Briefly, we
could define \textit{haplotypes} as a combination of allelic variants, each one
inherited from a parent. Instead, the \textit{genotype} is combined information
of the haplotypes. For example, humans have two haplotypes, being diploid, and
the genotype is their combined information. So, from 2005, publication of the
\textbf{GWAS} has begun. The goal of this type of studies is to screen the

@@

example in order to study outcomes of diseases.  In this particular historical
period, it is impossible not to mention viruses. In fact, by nature, viruses
replicate a lot but often in a not perfect way, during infections. Thus produces
many inexact clones, referred as \textbf{viral haplotypes}, which, taken
together, form the \textbf{viral pangenome}. The identification of all these
haplotypes is crucial both to the study of the spread of viruses and to the
production of efficient drugs, in a context of high pharmacological resistance.

@@

  (\textit{PBWT})}, proposed by Durbin in 2014 \cite{pbwt_durbin}. \textit{PBWT}
aims to represent efficiently an haplotypes panel, storing them in a queryable
compressed way (details will be described later). According to Durbin, the used
memory by \textit{PBWT} is nearly six times smaller than the raw data. Without
these spatial results it would has been impossible to process real information.
% \pb{qua dovresti essere più specifico sull'obiettivo di ridurre la memoria
  % per tenere gli aplotipi e fare analisi come l'estrazione di pattern
  % comuni...non hai menzionato le questioni della memoria...la compressione
  % serve a quello}
%\dc{I dettagli sono stati messi nello stato dell'arte}
Using this particular data structure, it is
possible to study only biallelic panel, over an alphabet $\Sigma\in\{0,1\}$, for
example, in order to extract common patterns in a set of haplotypes. Thanks to
the compressed representation of the 
data, the use of the \textit{PBWT} is found in many 
softwares for the study of haplotypes and in various genotype imputation methods
that infer unobserved genotypes in a sample of individuals \cite{impute5}.
%\pb{idem qua ...se citi anche il risparmio di memoria in sintesi può essere
%utile}
%\dc{Aggiunto il fattore di riduzione indicato da Durbin nel paper}
% During the development of my master thesis, I worked to create a run length
% encoded variant of the \textit{PBWT}, the \textbf{RLPBWT}, using and adapting
% the various theories developed for the \textit{RLBWT}, in collaboration with
% the authors of \textit{MONI} and \textit{PHONI}.
% \pb{di in sintesi cosa è l'idea del run length encoding e sopra direi cha la
% BWT ha rivoluzionato gli algoritmi e i tools usati in Bioinformatica perchè è
% possibile indicizzare il genoma e i genomi in poco spazio ...dire come si
% riduce lo spazio ad esempio se si passa a MONI}
% \dc{Aggiunta breve spiegazione al RL quando introdotto}
% \dc{Aggiunto risparmio di memoria di MONI rispetto a BWA-MEM}
% In this context, my PhD is going to be focused on the development of new
% algorithms in 
% various topics related to open problems in the study of the
% \textit{haplotyping/genotyping}, of the \textit{genome 
%   variants} and of the imputation issues related. My intention is also to
%   deepen 
% experimental themes  
% of \textit{pattern matching} and of the \textit{pangenome graph}, in detail
% the 
% new developments on \textit{BWT} and indexing structures, as well as
% \textit{succinct data structures}, with particular attention to the use of
% \textit{bitvectors}. PARTE MESSA A FINE INTRODUZIONE
%\pb{se rimandi a dopo allo stato dell'arte queste cose la mia paura è che non
%capisco questa parte...devi a mio parere antcipare l'indispensabile prima
%dicendo che i dettagli li dai nella sezione dello stato dell'arte} 
%\pb{lo stato dell'arte serve a dire cosa si è fatto prima e cosa tu pensi di
%fare per avanzare nello stato dell'arte...di solito serve ad evidenziare i
%problemi aperti} 

@@

Now I present a brief overview of the main algorithms, data structures and tools
that are the state of the art of computational pangenomics.  

@@

%\pb{questa parte io la imposterei con titoli diversi che evidenziano da quali}
%problemi aperti dello stato dell'arte parti per poi sviluppare gli obiettivi
%del progetto...ad esempio sui Bitvectors ...se metti invece come titolo
%qualcosa tipo: 
% Efficient and compact boolean data structures and applications...
% Inizierei dicendo che una struttura dati estremamente utile per algoritmi di
% analisi sequenze è quello dei bitvectors utilizzati tra l'altro anche per
% implementare Bloom filters...sono compatti etec... magari anche dire chi ha
% fatto SDSL in che anno ...} 
% \dc{Titoli modificati come indicato}
% \paragraph*{Bitvectors}
\paragraph*{Efficient and compact boolean data structures.}
% \dc{Non so se ho lo spazio per introdurre i bloom filters}
If $\mathcal{X}$ is the optimal number of bits needed to store some data, a
representation of this information is defined \textit{succint} if it takes
$\mathcal{X} +o(\mathcal{X})$ bits of space. 
\textbf{Bitvectors} are one of the most important \textit{succint data
  structure}.  

@@

integer $i$, such that $0<i\leq rank_B(n)$, we can define

@@

\textit{constant time}, with the additional cost of $o(n)$ bits in
memory. There are several implementations of the 
same (for example \textit{plain bitvector, interleaved
  bitvector, sparse bivector} etc$\ldots$) within \textbf{SDSL (\textit{Succint
    Data Structures Library})}, one of the most important C++ library used in
bioinformatics proposed by Gog et al. in 2014. Thanks to the various
implementations, both the computational time of the two main 
operations and the additional bits needed vary, allowing a
better choice of the best variant possible depending on the use case. Due to
their compactness in memory, \textit{bitvectors} are widely used in algorithms
for analysis of biological sequences. An example of the use of is 

@@

for indexing and mapping. 
%\paragraph*{RLBWT}
\paragraph*{Run length encoding and succinct data structures.}
% \pb{Run length encoding and succinct data structures... potrebbe essere il
% titolo generale qua...tieni presente che il run lengh encoding è una vecchia
% idea usata nella compressione dei dati alternativa a LZ.. Lempel Ziv...ed è
% stata estesa alla BWT...} 

@@

\textit{FM-index}.
Speaking of \textit{pangenome}, linear indexing via \textit{FM-index} is no
longer the 

@@

the \textbf{Run-Length encoded Burrows–Wheeler Transform
  (\textit{RLBWT})}. Given a text 

@@

\textbf{thresholds}, defined as the minimum \textit{LCP} value between two
consecutive runs of the same character, in \textit{MONI}. Instead, in
\textit{PHONI}, the authors use the \textbf{longest common extension
  (\textit{LCE}) query}, to compute the \textit{MS}. Formally, an \textit{LCE
  query}, given two position $i$ and $j$ in a text $T$, such that $|T|=n$,
compute the length of the longest common prefix between $T[i:n-1]$ and
$T[j:n-1]$, so an \textit{LCE} is the right equal common extension 
between two position in the text. Both solutions operate on a compressed
representation of the text, via a
\textbf{straight-line program (\textit{SLP})}. Briefly an \textit{SLP} is based
on a \textit{grammar-compression} algorithm, by a \textit{context-free grammar},
and here it is used for \textit{random access} and for \textit{LCE queries}. 

@@

extended either to the right or to the left. Furthermore, using a particular

@@

in $SA$, the previous and next index stored in the suffix array.

@@

So, it was possible to perform pattern matching

@@

a pangenomic context.
\paragraph*{PBWT.}

@@

iff $h$ is the smallest column index such that $y_i^k[h,k)=y_{i-1}^k[h,k)$. 

@@

research based on it, both in terms of variant design, such as the
\textbf{multiallelic PBWT} or the \textbf{dynamic PBWT}, where the static PBWT
is generalized to a dynamic data 
structure via linked lists, and in terms of its use to study haplotype panels.

@@

signatures of natural selection.
Another interesting paper is the one by Williams and Mumey, published in 2020,

@@

addition of a \textit{missing data} management attempt, handled with the help of
wildcards, using the \textit{PBWT}. 

@@

software making genotype imputation in order to predict unobserved genotypes
from a panel with milions of haplotypes, so it was necessary to use
\textit{PBWT}. Improving this data structure, it will be possibile to help the
\textit{GWAS} on tumors and other disorders. 
% Durbin himself, the author of the \textit{PBWT}, with Shchur et al. in 2019,
% proposed a use of his data structure in \textit{GWAS} context. In fact, they
% studied the associations between genetic variants in order to identify signals
% of natural selection and to build the so-called \textbf{ancestral recombination
%   graph}, that contains complete information about history of samples.
% For the sake of completeness, an example where \textit{PBWT} is not used is
% \textbf{Ranbow}, proposed by Moeinzadeh et al. in 2020. The 
% aim of this project was the haplotype reconstruction of polyploid genome from
% short read sequencing data, studying also the multi-allelic case.
% togliere ranbow e paper durbin

@@

For my master's thesis, I had to adapt the concept described for
\textit{RLBWT} in \textit{MONI} and \textit{PHONI}. 

@@

context.
Summarizing, the most important research goal of my PhD is designing and
validating algorithms and data structures to deal with open problems in
pangenomics, such as: 
\vspace{-1.25mm}

@@

  current use of bitvectors to handle efficiently also the \textit{LF-mapping}
  with more than two alleles. 

@@

  Handling missing data is known to be \textit{hard} because every gap could
  assume any possible value. Having said this, I should  
  probably deal with \textit{parametric} or \textit{approximate algorithms},
  based on researches already developed in \textit{BIAS}.
  \item \textbf{Algorithms for pangenome graph \& other goals}. Another research
  goal is to deepen 
  the study of \textit{pangenome graph}. In fact, it is possibile to interpret
  haplotype sequences as a 
  pangenome graph, in a compact way via the \textbf{Graph BWT (\textit{GBWT})},
  even if it is more a \textit{multi-string BWT} than a classical graph. A first
  improvement would be a spatial optimization. There are a lot of open problems 
  regarding the use of \textit{GBWT}, such as handling missing / erroneous data
  and representing complex and nested 
  variants. Once I have tried to solve the problem of missing data for the
  \textit{PBWT/RLPBWT}, I could think of an adaptation 
  for the \textit{GBWT}, because it would be possibile to
  analyze more precisely real data. I would also point out my interest in
  deepening all the other issues related to 
  pattern matching, indexing structures and pangenome graph, according to the
  \textit{BIAS} laboratory's topics of interest.
  % chiarire questa parte
  %\item \textbf{Imputation}.
  % As introduced, I have
  % developed a wide-range interest in the application of computer science to
  % biology. So, I do 
  % not intend to forget the actual use of the data structures that I will study
  % and develop during my PhD. The management of missing and
  % multiallelic data can further improve the state of the art of \textit{GWAS}
  % and of genotype imputation, allowing even more precise inference of unobserved
  % genotypes. All this must be read in the perspective of 
  % a continuous rise of available data, not only regarding the human
  % sequencing. \\ 

@@

\vspace{-1.25mm}
The theoretical research requires an experimental verification, for this purpose
some technologies that will be used during my PhD must be mentioned. 
During both my bachelor and master degree, I mainly focused on the use of the
\textbf{C++ programming language}, thanks to the
availability of efficient libraries, such as the already 
cited \textit{SDSL}. However, in

@@

\textit{computational systems biology}.  

@@

my master thesis, including Christina Boucher (University of Florida)
and Travis Gagie (Dalhousie University).
% \pb{nella bibliografia il lavoro MONI non è forse apparso a recomb22? non hai
% una referenza più recente?}
% \dc{non trovo la ref ufficiale a recomb21, eventualmente la posso scrivere a
% mano ma ho aggiornato con la ref a Journal of Computational Biology del 2022} 
% \pb{non metterei la referenza di Natural Computing...ma metterei quella di
% Briefings in bioinformatics sul Pangenome consortium che trovi in Natural
% computing..}
% \dc{non sono certo di aver trovato la reference indicata}
% \bibliographystyle{abbrv}
% \bibliography{abstract}
\printbibliography
% \newpage
% \noindent
% \begin{shaded}
%   \noindent
%   \textbf{Note}:
%   \begin{itemize}
%     \item il font dovrebbe essere corretto secondo le richieste. Non ho visto
%     specifiche relative ai margini del file
%     \item nell'introduzione mancano alcuni concetti relativi allo
%     studio diretto del grafo pangenomico. La professoressa Bonizzoni aveva
%     citato ad esempio le \textit{superbubbles}. Una volta chiariti gli scopri
%     precisi 
%     del progetto di ricerca dovrò sistemare (oltre che aggiungere eventuali
%     aspetti non considerati e toglierne altri)
%     \item nell'introduzione devo capire come ordinare meglio i concetti
%     \item nell'introduzione forse servono esempi di casi d'uso relativi alle
%     varie problematiche. Probabilmente è necessaria una parte più discorsiva
%     relativa anche all'importanza biologica di questo genere di studi
%     \item la conclusione dell'introduzione, dove si parla di cosa verrà
%     eventualmente approfondito durante il PhD, è praticamente un
%     \textit{placeholder}  
%     \item l'incipit allo stato dell'arte è sicuramente da modificare
%     \item la struttura dell'intera sezione riguardante lo stato dell'arte penso
%     debba essere rivista in base a standard per la stesura della proposal che
%     non conosco
%     \item nello stato dell'arte la parte relativa alla BWT è commentata in
%     quanto, pur avendola scritta, non la ritengo essenziale (soprattutto avendo
%     a che fare con un limite di 4 pagine)
%     \item nello stato dell'arte la sezione relativa ad alcuni esempi di studio
%     sugli aplotipi deve essere profondamente rivista in luce soprattutto dei
%     research goals. Un esempio potrebbe essere la citazione, ad esempio, di
%     HapCol, qualora i goals si spostassero un po' dalla PBWT e dai suoi usi più
%     immediati 
%     \item per ignoranza mia su come vada scritto questo tipo di documento non
%     sono sicuro che alcuni dettagli formali nei vari passaggi dello stato
%     dell'arte siano sensati da mettere. Stesso discorso vale nell'incipit dei
%     research goals
%     \item la citazione di quanto svolto per la tesi magistrale non so se sia
%     necessario, e, qualora lo fosse, se la collocazione all'inizio dei research
